---
title: "GP40 LSA"
author: "Yamé Meyvaert"
date: "2024-11-28"
output: html_document
---

## Loading the data

Installing and loading the libraries:

```{r}
#install.packages("tidyverse")
library(tidyverse)
library(readxl)

```

```{r}
cancer_dataset <- read_csv('metadata.csv')

cancer_abundances <- read_excel('abundances.xlsx')
```

###Data exploration Metadata

Checking datatypes:
```{r}
str(cancer_dataset)
```

Converting datatypes (cleaning the data)
```{r}
cancer_dataset$Gender <- as.factor(cancer_dataset$Gender)
cancer_dataset$Proteomic_Subtype <- as.factor(cancer_dataset$Proteomic_Subtype)
cancer_dataset$`Diameter_of_tumor_(cm)` <- as.numeric(cancer_dataset$`Diameter_of_tumor_(cm)`)
cancer_dataset$Survival_status <- as.factor(cancer_dataset$Survival_status)
cancer_dataset$Recurr_status <- as.factor(cancer_dataset$Recurr_status)

#checking for correct converting
str(cancer_dataset)
```

Check the header
```{r}
head(cancer_dataset)
```

Checking for missing values:
```{r}
missing_values <- colSums(is.na(cancer_dataset))
print(missing_values)
```

Interpretation: 7 missing values in Diameter_of_tumor_(cm) column and 10 in Tumor_cellularity_(%).

Missing data visualisation:
```{r}
#install.packages("naniar")
library(naniar)

vis_miss(cancer_dataset)       # Heatmap van missende waarden
gg_miss_var(cancer_dataset)    # Barplot van missende waarden per kolom
```

Basic stats for each of the columns:
```{r}
summary(cancer_dataset)
```

Distribution of the data
```{r}
cancer_dataset %>%
  summarise(across(where(is.numeric), list(mean = mean, sd = sd, min = min, max = max), na.rm = TRUE))
```

```{r}
cancer_dataset%>%
  count(Proteomic_Subtype)
```

###Data exploration Abundances data

```{r}
str(cancer_abundances)
```

Check the header
```{r}
head(cancer_abundances)
```

```{r}
cancer_abundances %>%
  mutate(missing_count = rowSums(is.na(.)))%>%
  select(ID, missing_count)
```

```{r}
summary(cancer_abundances)
```


Hypothese 1:
=> Heeft de proteomic subtype een correlatie met de Tumor cellularity:
Hypotheses:
-Null Hypothesis (H0): The median Tumor_cellularity_(%) is the same across all Proteomic_Subtype groups.
-Alternative Hypothesis (Ha): At least one group has a different median Tumor_cellularity_(%).

Missing values:
```{r}
vis_miss(cancer_dataset)       # Heatmap van missende waarden
gg_miss_var(cancer_dataset)    # Barplot van missende waarden per kolom
```
Check how many NA values in Tumor cellularity:
```{r}
sum(is.na(cancer_dataset$`Tumor_cellularity_(%)`))
```
Random? Or not random?
```{r}
table(is.na(cancer_dataset$Proteomic_Subtype), is.na(cancer_dataset$`Tumor_cellularity_(%)`))

# Contingency table for missingness
missing_table <- table(is.na(cancer_dataset$Proteomic_Subtype), is.na(cancer_dataset$`Tumor_cellularity_(%)`))

# Chi-squared test
chisq.test(missing_table)

```
INterpretation: the p-value is below 0.05 which means that the missingness of the values is not random and therefore we can't remove them from the column. 

```{r}
#install.packages("naniar")
library(naniar)

#Plot Missingness:

# Check missing data patterns
vis_miss(cancer_dataset)  # naniar package

```

Replace NA with median:
```{r}
# Calculate the median while ignoring NA values
median_value <- median(cancer_dataset$`Tumor_cellularity_(%)`, na.rm = TRUE)

# Replace NA values with the calculated median
cancer_dataset$`Tumor_cellularity_(%)`[is.na(cancer_dataset$`Tumor_cellularity_(%)`)] <- median_value

# Print the column to check if NAs were replaced
print(cancer_dataset$`Tumor_cellularity_(%)`)

```

Outliers eruitfilteren:
```{r}
ggplot(cancer_dataset, aes(x = Proteomic_Subtype, y = `Tumor_cellularity_(%)`, fill = Proteomic_Subtype)) +
  geom_boxplot(outlier.color = "purple", outlier.shape = 8) +
  theme_minimal() +
  labs(title = "Tumor Cellularity by Proteomic Subtype", y = "Tumor Cellularity (%)", x = "Proteomic Subtype")

```

```{r}
df <- cancer_dataset %>% filter(`Tumor_cellularity_(%)` < quantile(`Tumor_cellularity_(%)`, 0.99))
ggplot(cancer_dataset, aes(x = Proteomic_Subtype, y = `Tumor_cellularity_(%)`)) +
  geom_boxplot(fill = "lightpink") +
  theme_minimal()
```
 => there is no difference when filtered on 99% quantile. 


###Data visualisation

Histogram for numerical values
```{r}
ggplot(cancer_dataset, aes(x = `Tumor_cellularity_(%)`)) +
  geom_histogram(binwidth = 5, fill = "red", color = "black") +
  theme_minimal()
```

```{r}
cancer_dataset %>% ggplot(aes(x = `Tumor_cellularity_(%)` , fill = Proteomic_Subtype)) + geom_histogram() +theme_classic()
```

Boxplot:
```{r}
ggplot(cancer_dataset, aes(x = Proteomic_Subtype, y = `Tumor_cellularity_(%)`)) +
  geom_boxplot(fill = "lightgreen") +
  theme_minimal()
```
Normal distribution per Subtype:
```{r}
library(dplyr)

# Perform Shapiro-Wilk test for normality on each subtype
normality_results <- cancer_dataset %>%
  filter(!is.na(`Tumor_cellularity_(%)`)) %>%  # Remove rows with NA values
  group_by(Proteomic_Subtype) %>%             # Group by subtype
  summarise(
    shapiro_p_value = shapiro.test(`Tumor_cellularity_(%)`)$p.value
  )

print(normality_results)

```
A small p-value (< 0.05) suggests the data is not normally distributed.

Visualisation:
```{r}
ggplot(cancer_dataset, aes(x = `Tumor_cellularity_(%)`, fill = Proteomic_Subtype)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot of Tumor Cellularity by Subtype", 
       x = "Tumor Cellularity (%)",
       y = "Density") +
  theme_minimal()
```
Tumor cellullarity data are not normally ditributed per Proteomic Subtype. (Log-transformations are not an improvement, it was tested). 
ANOVA can not be performed, because the data is not normally distributed. 

P-values testing: 
```{r}
normCel <- cancer_dataset %>%
  group_by(Proteomic_Subtype) %>%
  summarize(p_value = t.test(`Tumor_cellularity_(%)`)$p.value)

# Print results
print(normCel)
```


Kruskal-Wallis test
```{r}
kruskal.test(`Tumor_cellularity_(%)` ~ Proteomic_Subtype, data = cancer_dataset)
```
Interpretation: With a p-value of 
0.01354, which is less than the common significance level (α=0.05), the test result is statistically significant.
This means that there is strong evidence to reject the null hypothesis.
At least one of the Proteomic_Subtype groups has a significantly different median Tumor_cellularity_(%).

Since the Kruskal-Wallis test is an omnibus test, it doesn't specify which groups differ. To identify the specific groups with differing medians, you can perform pairwise post-hoc tests: (post-hoc tests are needed to specify which groups differ)

```{r}
pairwise.wilcox.test(cancer_dataset$`Tumor_cellularity_(%)`, 
                     cancer_dataset$Proteomic_Subtype, 
                     p.adjust.method = "bonferroni")


#Visualisation:
ggplot(cancer_dataset, aes(x = Proteomic_Subtype, y = `Tumor_cellularity_(%)`, fill = Proteomic_Subtype)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Tumor Cellularity by Proteomic Subtype", y = "Tumor Cellularity (%)")

```
Interpretation: Results Interpretation
- S-I vs S-II:
Adjusted p-value = 1.000.
There is no statistically significant difference in the median Tumor_cellularity_(%) between subtypes S-I and S-II.

- S-I vs S-III:
Adjusted p-value = 0.059.
This is borderline significant at the 0.05 level. You might consider the difference to be marginal or weak evidence against the null hypothesis.

- S-II vs S-III:
Adjusted p-value = 0.026.
This indicates a statistically significant difference in the median Tumor_cellularity_(%) between subtypes S-II and S-III at the 0.05 level.

The warning occurs due to ties in Tumor_cellularity_(%), which is common with real-world data. This doesn't invalidate your results:

Monte Carlo approximation for more robust p-values:
```{r}
pairwise.wilcox.test(cancer_dataset$`Tumor_cellularity_(%)`, 
                     cancer_dataset$Proteomic_Subtype, 
                     p.adjust.method = "bonferroni", 
                     exact = FALSE)

```
With the Monte-Carlo approximation there is no difference in p-values. When ties are few and the data size is reasonable, the normal approximation (used in exact = FALSE) provides a good estimate of the p-values. Therefore thee approximate p-values provided by the pairwise Wilcox test are reliable and consistent with the results generated via Monte Carlo simulations.


The Wilcoxon test is less sensitive to outliers.

Hypothese 2:
=> Welke eiwitten hebben invloed op de survival status
==> 
```{r}

```

